How to select predictive models for decision-making or causal inference?

### 2023-12-16 how_to_select_causal_models.tex
iter

\begin{table*}[bt!]
    \makebox[\textwidth]{
        \begin{threeparttable}[b]
            \caption{Review of causal risks
                ---
                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
                \citet{schuler_comparison_2018}.
                \label{tab:evaluation_metrics}}
            \centering
            \begin{tabular}{llr}
                \toprule
                Risk                                                                                           & Equation
                                                                                                               & Reference                                                                                                                                           \\
                \midrule
                $mse(\tau(X), \tau_f(X))=\tau\text{-risk}$                                                     & $\mathbb E_{X\sim
                            p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
                                                                                                               & Eq. \ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
                $mse(Y, f(X)) = \mu\text{-risk}$                                                               & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
                                                                                                               & Def. \ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
                $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
                (Y-f(X ; A))^2 \right]$                                                                        & Def.
                \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
                $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
                Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
                \\
                $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
                \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
                Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
                \\
                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
                -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
                Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
                \\
                \bottomrule
            \end{tabular}
        \end{threeparttable}
    }
\end{table*}

@@

We use simulated data, on which the ground-truth causal effect is known. Going
beyond prior empirical studies of causal model selection
\cite{schuler_comparison_2018,alaa_validating_2019}, we use many
generative processes, to reach more general conclusions (as discussed
in \ref{apd:results:fig:seed_effect}).

We generate the response functions using random bases extension, a common method
in biostatistics, \eg functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, we control the complexity of the non-linear outcome
models. We use random approximation of Radial Basis Function (RBF) kernels
\cite{rahimi_random_2008} to generate the outcome and treatment functions. RBF
use the same process as polynomial splines but replace polynomial by Gaussian
kernels. Unlike polynomial, Gaussian kernels have decreasing influences in the
input space. This avoids unrealistic divergences of the response surfaces at the
ends of the feature space. We generate 1\,000 datasets based on these functions,
with random overlap parameters. Example shown in Figure
\ref{fig:simulation_examples} and details in \ref{apd:experiments:generation}.

@@

We compare different candidate estimators that approximate imperfectly the
data-generating process. To build such estimators, we first use a RBF
expansion similar to the one used for data generation. We choose two random
knots and apply a transformation of the raw data features with a Gaussian
kernel. This step is referred as the featurization. Then, we fit a linear
regression on these transformed features. We consider two ways of combining
these steps for outcome mode; we use common nomenclature
\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
meta-learners that differ on how they model, jointly or not, the treated and the
non treated:
\begin{itemize}
    \item SLearner: A single learner for both population, taking the treatment as
          a supplementary covariate.
    \item SftLearner: A single set of basis functions is sampled at random for both
          populations, leading to a given feature space used to model both the treated and
          controls, then two
          separate different regressors are fitted on this shared representation.
    \item TLearner: Two completely different learners for each population, hence
          separate feature representations and regressors.
\end{itemize}

@@

We also use three benchmarks of the causal-inference literature, semi-simulated
data adding a known synthetic causal effect to real --non synthetic-- healthcare
covariate. ACIC 2016 \cite{dorie_automated_2019} is based on the Collaborative
Perinatal Project \cite{niswander_women_1972}, a RCT studying infants’
developmental disorders containing 4,802 indivduals and 55 features. We used 770
dataset instances: 10 random seeds for each of the 77 simulated settings for the
treatment and outcomes.  ACIC 2018 \cite{shimoni_benchmarking_2018} simulated
treatment and outcomes for the Linked Births and Infant Deaths Database (LBIDD)
\cite{macdorman_infant_1998} with $D=177$ covariates. We used all 432 datasets
of size $N=5\,000$.  Twins \cite{louizos_causal_2017} is an augmentation of real
data on twin births and mortality rates \cite{almond_costs_2005}. There are
$N=11\,984$ samples, and $D=50$ covariates for which we simulated 1,000
different treatment allocations. Datasets details are given in Appendix
\ref{apd:experiments:datasets}.

@@

\subsection{Results: factors driving good model selection}\label{empirical_study:results}

@@

\paragraph{Selecting different seeds and parameters is crucial to draw
    conclusions}\label{{apd:results:seed_effect}}

>>>>

\begin{table*}[bt!]
    \makebox[\textwidth]{
        \begin{threeparttable}[b]
            \caption{Review of causal risks
                ---
                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
                \citet{schuler_comparison_2018}.
                \label{tab:evaluation_metrics}}
            \centering
            \begin{tabular}{llr}
                \toprule
                Risk                                                                                           & Equation
                                                                                                               & Reference                                                                                                                                           \\
                \midrule
                $\tau\text{-risk}=\text{MSE}(\tau(X), \tau_f(X))$                                                     & $\mathbb E_{X\sim
                            p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
                                                                                                               & Eq. \ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
                $\mu\text{-risk} = \text{MSE}(Y, f(X))$                                                               & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
                                                                                                               & Def. \ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
                $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
                (Y-f(X ; A))^2 \right]$                                                                        & Def.
                \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
                $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
                Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
                \\
                $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
                \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
                Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
                \\
                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
                -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
                Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
                \\
                \bottomrule
            \end{tabular}
        \end{threeparttable}
    }
\end{table*}

@@

We use simulated data, on which the ground-truth causal effect is known. Going
beyond prior empirical studies of causal model selection
\cite{schuler_comparison_2018,alaa_validating_2019}, we use many
generative processes, which is needed to reach general conclusions (Appendix \ref{apd:results:seed_effect}).

We generate the response functions using random bases extension, a common method
in biostatistics, \eg functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, we control the complexity of the non-linear outcome
models. We use random approximation of Radial Basis Function (RBF) kernels
\cite{rahimi_random_2008} to generate the outcome and treatment functions. RBF
use the same process as polynomial splines but replace polynomial by Gaussian
kernels. Unlike polynomial, Gaussian kernels have decreasing influences in the
input space. This avoids unrealistic divergences of the functions at the
ends of the feature space. We generate 1\,000 datasets based on these functions,
with random overlap parameters. Example shown in Figure
\ref{fig:simulation_examples} and details in \ref{apd:experiments:generation}.

@@

We test model selection across different candidate estimators that
approximate imperfectly the
data-generating process. To build such estimators, we first use a RBF
expansion similar to the one used for data generation. We choose two random
knots and apply a transformation of the raw data features with a Gaussian
kernel. This step is referred as the featurization. Then, we fit a linear
regression on these transformed features. We consider two ways of combining
these steps for outcome mode; we use common nomenclature
\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
meta-learners that differ on how they model, jointly or not, the treated and the
non treated:
\begin{itemize}
    \item SLearner: A single learner for both population, taking the treatment as
          a supplementary covariate.
    \item SftLearner: A single set of basis functions is sampled at random for both
          populations, leading to a given feature space used to model both the treated and
          controls, then two
          separate different regressors are fitted on this shared representation.
    \item TLearner: Two completely different learners for each population, hence
          separate feature representations and regressors.
\end{itemize}

@@

We also use three semi-simulated
data adding a known synthetic causal effect to real --non synthetic-- healthcare
covariate. ACIC 2016 \cite{dorie_automated_2019} is based on the Collaborative
Perinatal Project \cite{niswander_women_1972}, a RCT studying infants’
developmental disorders containing 4,802 indivduals and 55 features. We used 770
dataset instances: 10 random seeds for each of the 77 simulated settings for the
treatment and outcomes.  ACIC 2018 \cite{shimoni_benchmarking_2018} simulated
treatment and outcomes for the Linked Births and Infant Deaths Database (LBIDD)
\cite{macdorman_infant_1998} with $D=177$ covariates. We used all 432 datasets
of size $N=5\,000$.  Twins \cite{louizos_causal_2017} is an augmentation of real
data on twin births and mortality rates \cite{almond_costs_2005}. There are
$N=11\,984$ samples, and $D=50$ covariates for which we simulated 1,000
different treatment allocations. Datasets details are given in Appendix
\ref{apd:experiments:datasets}.

@@

\section{Results: factors driving good model selection}\label{empirical_study:results}

@@

\paragraph{Selecting different seeds and parameters is crucial to draw
    conclusions}\label{apd:results:seed_effect}
